\pdfbookmark{Acknowledgments}{Acknowledgments}
\begin{acknowledgments}

    This thesis would not have been possible without the support of
    many. Family and friends have supported me over
    the years to various degrees.
    
    First of all, I dedicate this thesis to my mother, Maria da
    Conceição, and my father, Júlio António. I profusely thank them
    for everything they have done in my education, both academically
    and personally. I especially thank them for supporting me
    immediately after I told them that I wanted to pursue a Master's
    degree in Artificial Intelligence in Edinburgh, even though I had
    already started one in Biomedical Engineering. Through a domino
    effect, this has culminated in this thesis, and I hope I'm making
    them proud. I also want to thank my grandmother, Maria Helena, as she
    always cared for me as though I were her child. I'm also grateful
    to the rest of my family for their support and encouragement.
    
    This thesis would have been much harder to do, had I not had
    endless support from Sofia. She has been a pillar in my life
    throughout these years, and I'm glad to have her as my partner.
    Many ideas have blossomed from conversations with her over dinner
    or on the long walks we usually take. Moreover, she has drawn
    some of the figures present in this thesis (the reader can infer
    which ones by how much better they look).
    
    I am forever grateful to my official advisors, André and Vlad,
    and my honorary (but just as important) advisor, Wilker. They
    have been stellar, and I am fortunate to consider them not only
    advisors but also friends. I thank André for believing in me in
    the first place and inviting me to start on this journey with
    him. I thank Vlad for always looking out for me in numerous ways
    and inspiring me to be better. I thank Wilker for introducing me
    to latent and variational modeling, topics I was and am highly
    interested in, and for being kind and patient while explaining
    those concepts.
    
    I am fortunate to have had the help and support of many friends.
    From school, I thank João Miguel, Ricardo B., Martim, Sérgio,
    Álvaro, Inês, Alice, Leonor, Miguel, Filipe, Ricardo F., Renato,
    and Luciano. I've known many of them since I was six years old,
    and I hope to have you by my side for countless more years! From
    IST (2012-2016), I thank Catarina, Patrícia, Mariana, Maria,
    Miguel, Joana, Diogo, Gonçalo, André Melo, Daiane, and João. They
    are incredible friends and an endless source of laughs,
    inspiration, and kindness. From Edinburgh, I thank Alasdair,
    Antonio, Lasse, Diogo, and Miguel. We all shared great
    discussions and experiences over there, learned from among the
    best, and despite some physical distance right now, I still feel
    like we are close. I also thank my \textit{sis-law} Inês, for
    being the little sister I never had, and my \textit{bro-law}
    Francisco, who would undoubtedly find it extremely funny to be
    mentioned in a doctoral dissertation. A special thanks belongs to
    João Miguel, André Melo, and João Graça. When I was most confused
    and lost in my academic journey, they were the ones who persuaded
    me (thankfully!) to change paths and pursue a career in Machine
    Learning. This thesis exists, in a way, because of conversations
    with them.
    
    Finally, I'm grateful for my colleagues at the SARDINE and
    Probabll labs, who I'm glad to call friends. I thank Ben, Erick,
    Tsveti, Marcos, Pedro, Chryssa, Taya, Nuno, Patrick, António,
    Bryan, and Lina. This endeavor would have been so dull had not
    all of them been around, and countless discussions led to many
    ideas present in this thesis. Additionally, even though we have
    never met in person, I thank Mathieu Blondel, who had the idea to
    learn $\alpha$ in $\alpha$-\entmaxtext
    (\chapref{chap:adaptsparse}), and had done initial work on
    top-$k$ sparsemax (\chapref{chap:sparsemarg}).
    
    My PhD was funded by the European Research Council (ERC
    StG DeepSPIN 758969).
    
\end{acknowledgments}
\clearpage
\thispagestyle{empty}
\cleardoublepage
